\LARGE \bf Planning with Attitude

### 2020-10-14 paper/main.tex
Update timing results

    The Jacobian of $r(q)$ can be found using \eqref{eq:quat_gradient}:
    \begin{align} \label{eq:newton_foc}
            \nabla r(q) &= \pdv{r}{q} G(q) \\
              &= -2 H^T R(q)^T \left( \sum_i R({}^B \hat{v}_i) \right) G(q) .
    \end{align}
    Given a guess solution, $q_k$, The standard Gauss-Newton method can then be used to compute a three-parameter update, $\phi_k$ via the Moore-Penrose psuedoinverse:

    \begin{equation}
    	\phi_k = ({\nabla r}^T {\nabla r})^{-1} {\nabla r}^T r(q_k).
    \end{equation}
	The update is then applied using the composition for the group:
	\begin{equation}
		\q_{k+1} = \q_k \otimes \varphi(\phi_k)
	\end{equation}.
	This ``multiplicative'' Gauss-Newton method is summarized in Algorithm \ref{alg:mgn}.

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrel roll, a quadrotor flip, and a
    satellite with flexible solar panels that must slew to a new orientation while
    avoiding a keep-out zone.    
    All problems are run using ALTRO, first without any of the modifications presented in
    the current paper, analagous to the naive Newton's method in section \ref{sec:Wahbas}
    and labeled ``naive'', and then using the modifications listed in Sec.
    \ref{sec:trajopt} and the geodesic cost function described in Sec.
    \ref{sec:cost_functions}, labeled ``modified''. All cost functions were of the
    following form:
    \begin{align}
        \ell_\text{naive}(x, \bar{x}, Q) &= \half (x - \bar{x})^T Q (x - \bar{x}) \\
        \ell_\text{modified}(x, \bar{x}, Q) &= \half (x-\bar{x})^T \bar{Q} (x-\bar{x}) + w (1
            \pm \bar{q}^T q)
    \end{align}
    where $\bar{x}$ is the reference state and $\bar{Q} = \text{diag}(Q_r, \vec{0}_4,
    Q_v, Q_\omega)$, with $Q_r,Q_v,Q_\omega$ being the weights of $Q$ for position, and
    linear and angular velocity, respectively.

    Timing results are summarized in Table \ref{tab:timing_results}. All experiments were
    solved to a constraint satisfaction tolerance of $10^{-5}$ and discretized with a 3rd
    order Runge-Kutta integrator. All results were run on a desktop computer with an AMD
    Ryzen 2950x processor with 40 GB of RAM. Code for all experiments is available on
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        A 180 degree barrel roll trajectory for a fixed-wing airplane was optimized. The
        airplane's dynamics model consists of the a simple rigid body as defined in
        Section \ref{sec:rigidbody_dynamics} with forces and torques due to lift and drag
        fit from wind tunnel data \cite{manchester2016udp}. The airplane was tasked to do
        a barrel roll by constraining the terminal state to upside-down (see Fig.
        \ref{fig:barrellroll}). To mitigate issues with integration error and drift in the magnitude of the quaternion,
        the following constraint function was used to enforce a terminal orientation of $\bar{q}$:
        \begin{equation}
            \frac{q_v}{\norm{q}} - \bar{q}_v
                \sign\left(\bar{q}^T \frac{q }{ \norm{q}}\right) = 0
        \end{equation}

        The solver was initialized with level flight trim
        conditions. The convergence of the different versions of ALTRO is compared in
        Fig. \ref{fig:c_max_convergence}. As expected, the modified version achieves
        better convergence and faster solve times compared to the na\"ive version since
        the expansions being provided to the algorithm more accurately capture the
        relationship between the attitude state and the goal and constraints. For this 
        relatively simple problem, we gained a modest 7\% improvement in runtime, despite
        the additional matrix multiplications when calculating the cost and constraint
        expansions.

        \begin{figure}[ht]
            \centering
            \includegraphics[width=\columnwidth]{figures/barrellroll.png}
            \caption{Barrel roll trajectory computed by ALTRO using a terminal cost to encourage an upside-down attitude.}
            \label{fig:barrellroll}
        \end{figure}

        \begin{figure}[ht]
            \centering
            \includegraphics[height=4cm,width=\columnwidth]{figures/c_max_convergence.tikz}
            \caption{Constraint satisfaction as a function of iteration when solving the barrel roll problem using ALTRO both with and without the modifications for optimizing unit quaternions.}
            \label{fig:c_max_convergence}
        \end{figure}

    \subsection{Quadrotor Flip}
        A 360 degree flip trajectory for a quadrotor was optimized with dynamics adapted
        from \cite{mellinger2012trajectory}. To encourage the flip, we specified a 
        ``waypoint'' cost function of the following form:
        \begin{multline}
                  \sum_{k \in \mathcal{N}} \ell(x_k, \hat{x}, \hat{Q}) 
                + \sum_{k \in \mathcal{W}} \ell(x_k, \bar{x}_k, Q_w) 
                + \half \sum_{k=1} u_k^T R u_k
        \end{multline}
        with $\mathcal{W} = \{45,51,55,75,101\}$, $\mathcal{N} = \{1:101\} \setminus
        \mathcal{W}$. The four intermediary ``waypoints'' were used to encourage the
        quadrotor to reach angles of \ang{90}, \ang{180}, \ang{270}, and \ang{360} around
        an approximately circular arc. The last waypoint was used to encourage the
        quadrotor to move towards the final goal. The quadrotor was constrained to stay
        above the floor and move to a goal state two meters away in the $+y$ direction. The
        solver was provided a dynamically infeasible initial trajectory that linearly
        interpolates between the initial and final states, rotating the quadrotor around
        the x-axis a full \ang{360}.

@@

\section{Conclusions} \label{sec:conclusion}
    We have presented a general, unified method for optimization-based planning and control for rigid-body
    systems with arbitrary attitude using standard linear algebra and vector calculus.
    The application of this methodology is straightforward and
    yields substantial improvements in the convergence of Newton and DDP-based methods, while also offering improvements for nonlinear constrained
    trajectory optimization for floating-base systems.

    Many state-of-the-art trajectory optimization methods, including direct collocation and sequential convex programming, rely on general-purpose optimization solvers whose internal numerical methods are not exposed to the user. Therefore, these methods are unable to exploit the full structure of both the trajectory optimization problem and the rotation group at a low level. In contrast, we are able to implement deep, native support for quaternions into the ALTRO solver, making it possible to solve more challenging problems with higher performance than other algorithms.

    In future work, we plan to apply ALTRO to real-time model-predictive control problems for aerial vehicles like quadrotors and airships that experience large attitude changes. The methods we have presented can also be leveraged to adapt other classes of gradient or Newton-based algorithms to exploit the structure of 3D rotations. Future directions beyond trajectory optimization may include simulation and planning methods that leverage maximal-coordinate formulations of multi-body dynamics~\cite{brudigam2020linear}, system-identification for complex multi-body systems and fixed-wing aircraft in post-stall conditions, and state estimation for spacecraft with sparse measurements.

>>>>

    The Jacobian of $r(q)$ can be found using \eqref{eq:quat_gradient}:
    \begin{equation} \label{eq:newton_foc}
            \nabla r(q) = \pdv{r}{q} G(q)
    \end{equation}
    \begin{equation}
            \nabla r_i(q) = -2 H^T R(q)^T R(\inframe{B}{\hat{v}_i}) G(q)
    \end{equation}
    Given a guess solution, $q_k$, The standard Gauss-Newton method can then be used to compute a three-parameter update, $\phi_k$ via the Moore-Penrose psuedoinverse:

    \begin{equation}
    	\phi_k = ({\nabla r}^T {\nabla r})^{-1} {\nabla r}^T r(q_k).
    \end{equation}
	The update is then applied using the composition for the group:
	\begin{equation}
		\q_{k+1} = \q_k \otimes \varphi(\phi_k).
	\end{equation}
	This ``multiplicative'' Gauss-Newton method is summarized in Algorithm \ref{alg:mgn}.

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrel roll, a quadrotor flip, and a
    satellite with flexible solar panels that must slew to a new orientation while
    avoiding a keep-out zone.    
    All problems are run using ALTRO, first without any of the modifications presented in
    the current paper, analagous to the na\"ive Newton's method in section \ref{sec:Wahbas}
    and labeled ``naive'', and then using the modifications listed in Sec.
    \ref{sec:trajopt} and the geodesic cost function described in Sec.
    \ref{sec:cost_functions}, labeled ``modified''. All cost functions are of the
    following form:
    \begin{align}
        \ell_\text{naive}(x, u, \bar{x}, Q, R) &= \half (x - \bar{x})^T Q (x - \bar{x}) + \half u^T R u \\
        \ell_\text{modified}(x, u, \bar{x}, Q, R) 
            &= \ell_\text{naive}(x, u, \bar{x}, \bar{Q}, R) + w (1 \pm \bar{q}^T q)
    \end{align}
    where $\bar{x}$ is the reference state and $\bar{Q} = \text{diag}(Q_r, \vec{0}_4,
    Q_v, Q_\omega)$, with $Q_r,Q_v,Q_\omega$ being the weights of $Q$ for position, and
    linear and angular velocity, respectively.

    Timing results are summarized in Table \ref{tab:timing_results}. All experiments are 
    solved to a constraint satisfaction tolerance of $10^{-5}$ and discretized with a 4th 
    order Runge-Kutta integrator. The results were run on a desktop computer with an AMD
    Ryzen 2950x processor with 40 GB of RAM. Code for all experiments is available on
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        A 180 degree barrel roll trajectory for a fixed-wing airplane was optimized. The
        airplane's dynamics model consists of the a simple rigid body as defined in
        Section \ref{sec:rigidbody_dynamics} with forces and torques due to lift and drag
        fit from wind tunnel data \cite{manchester2016udp}. The airplane was tasked to do
        a barrel roll by constraining the terminal state to be upside-down (see Fig.
        \ref{fig:barrellroll}). To avoid issues with the double-cover and integration error,
        the following was used to constrain a terminal orientation to $\bar{q}$:
        \begin{equation}
            \frac{q_v}{\norm{q}} - \bar{q}_v
                \sign\left(\bar{q}^T \frac{q }{ \norm{q}}\right) = 0
        \end{equation}

        The solver was initialized with level flight trim
        conditions. The convergence of the different versions of ALTRO is compared in
        Fig. \ref{fig:c_max_convergence}. As expected, the modified version achieves
        better convergence and faster solve times compared to the na\"ive version since
        the expansions being provided to the algorithm more accurately capture the
        relationship between the attitude state and the goal and constraints. For this 
        relatively simple problem, we gained a modest 11\% improvement in runtime, despite
        the additional matrix multiplications when calculating the cost and constraint
        expansions.

        \begin{figure}[t]
            \centering
            \includegraphics[width=\columnwidth]{figures/barrellroll.png}
            \caption{Barrel roll trajectory computed by ALTRO using a terminal cost to encourage an upside-down attitude.}
            \label{fig:barrellroll}
        \end{figure}

        \begin{figure}[t]
            \centering
            \includegraphics[height=4cm,width=\columnwidth]{figures/c_max_convergence.tikz}
            \caption{Constraint satisfaction as a function of iteration when solving the barrel roll problem using ALTRO both with and without the modifications for optimizing unit quaternions.}
            \label{fig:c_max_convergence}
        \end{figure}

    \subsection{Quadrotor Flip}
        A 360 degree flip trajectory for a quadrotor was optimized with dynamics adapted
        from \cite{mellinger2012trajectory}. To encourage the flip, we specified a 
        ``waypoint'' cost function of the following form:
        \begin{multline}
                  \sum_{k \in \mathcal{N}} \ell(x_k, u_k, \hat{x}, \hat{Q}, R) 
                + \sum_{k \in \mathcal{W}} \ell(x_k, u_k, \bar{x}_k, Q_w, R) 
        \end{multline}
        where $\hat{x}$, $\hat{Q}$ are the nominal state and state weight matrix, $Q_w$ is 
        the weight matrix for the waypoints, and
        $\mathcal{W} = \{45,51,55,75,101\}$, $\mathcal{N} = \{1:101\} \setminus
        \mathcal{W}$.
        Four intermediary ``waypoints'' were used to encourage the
        quadrotor to reach angles of \ang{90}, \ang{180}, \ang{270}, and \ang{360} around
        an approximately circular arc. The last waypoint was used to encourage the
        quadrotor to move towards the final goal. The quadrotor was constrained to stay
        above the floor and move to a goal state two meters away in the $+y$ direction. The
        solver was provided a dynamically infeasible initial trajectory that linearly
        interpolates between the initial and final states, rotating the quadrotor around
        the x-axis a full \ang{360}.

@@

\section{Conclusions} \label{sec:conclusion}
    We have presented a general, unified method for optimization-based planning and control for rigid-body
    systems with arbitrary attitude using standard linear algebra and vector calculus.
    We have demonstrated that the application of this methodology is straightforward and
    yields substantial improvements in the convergence of Newton-based methods (see Fig. 
    \ref{fig:wahba_convergence}), while also offering improvements for nonlinear constrained
    trajectory optimization for floating-base systems (see Table \ref{tab:timing_results}).

    \todo{This is a really nice pitch. Can we say something like this in the intro?}
    With the modifications presented, ALTRO can solve problems few other methods for
    trajectory optimization can. Many state-of-the-art methods, such as direct
    collocation or sequential convex programming, rely on commercial, proprietary, or
    general-purpose solvers whose internal numerics often are abstracted away from the
    user. By exploiting the unique structure of both the trajectory optimization problem
    and the rotation group, ALTRO will likely be able to solve more challenging problems
    with performance.

    The methods presented here can easily be leveraged to adapt other classes
    of gradient or Newton-based algorithms to exploit the structure of 3D rotations. Future
    work may include adaptation of methods for state estimation, localization, design, or 
    other methods for motion planning such as direct collocation.

    Future work will focus on continued 
    refinement of ALTRO and performance improvements for use in real-time model-predictive control applications, as well as extensions to multi-body robotic
    systems, such as humanoids or quadrupeds, represented in
    ``maximal'' coordinates that include the 3D orientation of each body \cite{brudigam2020linear}.
